---
title: "Testing"
author: "Marliese Baer"
format: html
editor: visual
---

Took the following code from data exploration where it isn't needed right now. If need it later, here it is.

### Ignore below (this section) for now

```{r}
CM |>
  mutate(Date = dmy(Date))
|>
  mutate(n_spp = as.double(n_spp)) |>
  mutate(Species = as.factor(Species)) |>
  mutate(Duration = as.double(`Duration (min)`)) |>
  select(Date, Species, n_spp, total_n, duration_min_day)

try |>
  group_by(Date, Species) |>
  summarise(spp_per_min = total_n/duration_min_day)

##why multiple spp_per_min values for each date and spp?

try |>
  mutate(Date = dmy(`Date `)) |>
  mutate(n_spp = as.double(n_spp)) |>
  mutate(Species = as.factor(Species)) |>
  mutate(Duration = as.double(`Duration (min)`))



Cape_may_clean <- Cape_May_2 |>
  pivot_longer(c(6:30), names_to = "Species_age",
               values_to = "n_spp") |>
  separate(Species_age, c("Species", "Age"),
           sep = " ") |>
  filter(Species == "RT" | Species == "SS" | Species == "CH" | Species == "AK" | Species == "PG") |>
  mutate(Date = dmy(`Date `)) |>
  mutate(n_spp = as.double(n_spp)) |>
  mutate(Species = as.factor(Species)) |>
  group_by(Date, Species) |>
  mutate(total_n = sum(n_spp)) |>
  mutate(Duration = as.double(`Duration (min)`))|>
  group_by(Date, Species) |>
  mutate(duration_min_day = sum(Duration)) |>
  select(Date, everything())


####after this point, I am doing work that should be in 'Data Exploration.qmd' but isn't because for some reason the date is back to a chr instead of date

try <- Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day) |>
  group_by(Date, Species) |>
  summarise(spp_per_min = mean(total_n/duration_min_day))

ggplot(data = try, aes(x = Date, y = spp_per_min)) +
  geom_point(aes(colour = Date)) +
  facet_wrap(~Species)

try2 <- Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day) |>
  group_by(Date, Species) |>
  mutate(spp_per_min = total_n/duration_min_day)

##add to below

Cape_may_clean$year <- lubridate::year(Cape_may_clean$Date)
Cape_may_clean$month <- lubridate::month(Cape_may_clean$Date)
Cape_may_clean$julian <- lubridate::yday(Cape_may_clean$Date) 

##when I run this code separately (not as a part of a pipe), it works. Why can't work in pipe

#need to convert to 3 columns: month, day, and year.
#plot per spp, spp_per_min for each month and year - so group by year and then plot x = month and y = spp_per_min

##julian date = x, spp/min = y, color = year, facet = spp <- spaghetti plot?

ggplot(data = try2, aes(x = ))

Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day) |>
  filter(Species == "RT") |>
  group_by(Date) |>
  summarise(spp_per_min = total_n/duration_min_day)

```

#### cape may stuff I think I can delete but am not ready to yet

trying to read in the original dataset and not the one that I deleted the first few columns of because there is a much much better way to do it in R as demonstrated below. Until can figure out how to do that, moving forward with the data I can easily access.

```{r}
####after this point, I am doing work that should be in 'Data Exploration.qmd' but isn't because for some reason the column names have a . at the end after I export

#make year, month, and julian date columns
#plot per spp, spp_per_min for each month and year - so group by year and then plot x = month and y = spp_per_min

##julian date = x, spp/min = y, color = year, facet = spp <- spaghetti plot?
Cape_may_clean$year <- lubridate::year(Cape_may_clean$Date)
Cape_may_clean$month <- lubridate::month(Cape_may_clean$Date)
Cape_may_clean$julian <- lubridate::yday(Cape_may_clean$Date) 


try <- Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day) |>
  group_by(Date, Species) |>
  summarise(spp_per_min = mean(total_n/duration_min_day))

ggplot(data = try, aes(x = Date, y = spp_per_min)) +
  geom_point() +
  facet_wrap(~Species)


##julian date = x, spp/min = y, color = year, facet = spp <- spaghetti plot?

try2 <- Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day, month, year, julian) |>
  group_by(Date, Species) |>
  mutate(spp_per_min = mean(total_n/duration_min_day))

ggplot(data = try2, aes(x = julian,
                        y = spp_per_min)) +
  geom_line(aes(colour = year)) +
  facet_wrap(~Species) +
  geom_smooth()

#above code: there are crazy outliers. How do I address/remove these?

#plot per spp, spp_per_min for each month and year - so group by year and then plot x = month and y = spp_per_min

try3 <- Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day, month, year, julian) |>
  group_by(Date, Species) |>
  mutate(spp_per_min = mean(total_n/duration_min_day)) |>
  group_by(year, month, Species) |>
  summarise(test = mean(spp_per_min))
           
ggplot(data = try3, aes(x = month,
                        y = test)) +
  geom_() +
  facet_wrap(~Species)

##how to fix graph above?

try4 <- Cape_may_clean |>
  select(Date, Species, n_spp, total_n, duration_min_day, month, year, julian) |>
  filter(Species == "RT") |>
  group_by(Date) |>
  mutate(spp_per_min = total_n/duration_min_day)

ggplot(data = try4, aes(x = julian,
                        y = spp_per_min)) +
  geom_point(aes(colour = year)) +
  geom_smooth()

ggplot(data = try4, aes(x = julian,
                        y = spp_per_min)) +
  geom_line(aes(colour = year)) +
  geom_smooth()


##what is the issue here (above)?
```

